---
title: "Unveiling the Spiritual Connection to Health"
author: "Mariela Mendez, Hannah Huang"
date: "2024-07-25"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

### Complementary and Alternative Medical (CAM) and Spirituality, Religiosity Survey
The CAM and Spirituality, Religiosity survey was given to chronically ill patients who were randomly chosen from a patient database at a private medical clinic in Southampton, UK. To be a participant for the survey, individuals were required to (1) have recent treatment within the last three years, (2) have a diagnosis of either migraines, irritable bowel, or chronic fatigue syndrome, and (3) have an age greater than 18. Six hundred questionnaires were originally mailed to the patients, yet there was a response rate of 18.5%, with 111 successfully completed questionnaires. The survey aims to explore the impact of spirituality on health. The spiritual topics range from spiritual experiences, personal values, to private religious practices and more. As for the health topics, these range from mindfulness practices to current health issues. The following data is sourced from the Association of Religion Data Archives, www.theARDA.com, and was collected by principal investigators Niko Kohls, Harald Walach, and George Lewish. The original data set consists of 254 variables and 111 participants, but for better comprehension, the ARDA decided to add variables and responses to include 258 total variables and 130 respondents. However, for the purposes of this analysis, the variable count has been reduced to 13 variables. While this is a much shorter data set, the chosen variables create a more direct approach. A concise data set facilitates clearer interpretation of results without one being overwhelmed by excessive data. 

### Motivation
Understanding the relationship between spirituality and health is an important aspect in well-being. In medical environments, especially those dominated by Western medicine, spirituality's impact on health is often disregarded. With this data set, investigators are able to explore how spirituality affects various sectors of health and overall well-being. Via analysis, patterns or correlations can be linked. One could uncover whether certain spiritual practices or beliefs are associated with better mental health, reduced stress, or improved coping mechanisms. In return, these patterns are relevant to healthcare providers or therapists looking to integrate effective strategies into their treatments. Notably, findings could influence policies related to holistic healthcare approaches that acknowledge the role of spirituality in patient well-being.

### Defining Variables
The original data set consists of 130 responses to 258 variables. Narrowing the focus for a more direct approach, 13 variables were chosen instead. Three categories were made in order to create a distinction between these chosen variables. The categories are as follows: spirituality/religious, health effects, and personal data. 

* For the **spiritual and religious** category, the following variables are as defined: 
    * *belgdwth* refers to the survey question "I believe in a God who watches over me",
    * *prvtpry* refers to "How often do you pray privately in place?",
    * *relpersn* refers to "To what extent do you consider yourself a religious person?"
    * *sprtprsn* refers to "To what extent do you consider yourself a spiritual person?". 
* The **health effects category** contains both physical and psychological effects. These variables are:
    * *DISTRS11*: poor appetite,
    * *DISTRS23*: nausea or upset stomach, 
    * *DISTRS37*: feeling weak in parts of your body, 
    * *DISTRS04*: feelings others are to blame for most of your troubles,
    * *DISTRS16*: feeling lonely,
    * *DISTRS53*: the idea that something is wrong with your mind. 
* As for the personal data category, the variables chosen are:
    * *gender*: female or male, 
    * *mrtlstus*: marital status (single, married, divorced,  or widowed), 
    * and *denom*: denomination. 


#### Rationale for Selecting Variables
Given the abundant amount of original variables, the elected variables were selected carefully. The variables *belgwth*, *prvtpry*, *relpersn*, and *sprtprsn* are variables that were chosen to determine how spiritual or religious a participant might be: 

* The variable *belgwth* indirectly questions the participant if they believe in a greater existence and if the participant believes that greater existence is watching them. 
* The variable *prvtpry* aims to see how religious a participant actually is outside of church. Oftentimes, individuals want a certain image of themselves to be displayed to the public, but private practices are a different topic.
* The variables *relpersn* and *sprtprsn* are a self ranking and sets how the participant views themselves on a scale from "not at all" to "very". 

The next set of variables relate to a participant's current health issues. They can be further split into physical and psychological effects. 

* The physical symptom variables were poor appetite (*DISTRS11*), nausea or upset stomach (*DISTRS23*), and feelings of weakness in one's body (*DISTRS37*). These variables were selected as they directly related to the requirements to the survey study of having irritable bowel or chronic fatigue syndrome. While the third indicator was migraines, the statements related to them were not as direct as these. So for the sake of a focused analysis, irritable bowel and chronic fatigue syndrome were preferred. 
* The psychological health effects chosen were feelings others were to blame for most of one's troubles (*DISTRS04*), feelings of loneliness (*DISTRS16*), and feelings that something is wrong with one's mind (*DISTRS53*). *DISTRS04* was chosen to see if participants would blame their problems on an entity rather than themselves while *DISTRS16* and *DISTRS53* were elected to see if the level of religiosity or spirituality affected how lonely one was or if they felt something was genuinely wrong with them. 

The last set of variables were demographic variables to get to know the data set population. These variables include gender, marital status, and denomination. Ultimately, these variables were chosen to illuminate any findings on the health relationships between the spiritual and physical in hopes to offer a shorter but to some extent holistic approach. 

(While creating our new dataset, we renamed certain variables for clarity during data analysis as follows: 

* *belgdwth* to *godwatches*
* *DISTRS11* to *poor_appetite*,
* *DISTRS23* to *nausea*, 
* *DISTRS37* to *weakness*, 
* *DISTRS04* to *blameothers*,
* *DISTRS16* to *lonely*,
* *DISTRS53* to *mindwrong*.)


```{r}
#load data set
load("/Users/marielamendez-pimentel/Documents/EDA Notebook/Datasets/CAM Original Dataset.RData")
#create new name for data set
camdata <- Complementary_and_Alternative_Medical_CAM_and_Spirituality_Religiosity_Survey
```

```{r}
#library access
##install.packages(tidyverse)
##install.packages(haven)
##install.packages(purrr)
##install.packages(knitr)
library(tidyverse)
library(haven)
library(purrr)
library(knitr)

#variables are stored as labeled vectors in the original data set so it must be converted to a regular vector so commands like summary apply to the variable
#create new  dataset, camdata_clean
camdata_clean <- camdata %>% 
  mutate(belgdwth = as_factor(belgdwth), #mutate labeled vectors to factors
         prvtpry = as_factor(prvtpry),
         relpersn = as_factor(relpersn),
         sprtprsn = as_factor(sprtprsn),
         DISTRS04 = as_factor(DISTRS04),
         DISTRS11 = as_factor(DISTRS11),
         DISTRS16 = as_factor(DISTRS16),
         DISTRS23 = as_factor(DISTRS23),
         DISTRS37 = as_factor(DISTRS37),
         DISTRS53 = as_factor(DISTRS53),
         gender = as_factor(gender),
         mrtlstus = as_factor(mrtlstus),
         denom = as_factor(denom)
        ) %>%
  select(belgdwth, prvtpry, relpersn, sprtprsn, DISTRS04, DISTRS11, DISTRS16, DISTRS23, DISTRS37, DISTRS53, gender, mrtlstus, denom) %>% #select variables of interest
  rename(godwatches = belgdwth, #rename for clarity
         poor_appetite = DISTRS11,
         nausea = DISTRS23,
         weakness = DISTRS37,
         blameothers = DISTRS04,
         lonely = DISTRS16,
         mindwrong = DISTRS53)

# summary(camdata_clean)

#save updated dataframe to github
save(camdata_clean, file = "/Users/marielamendez-pimentel/Documents/EDA Notebook/Datasets/camdata_clean.RData")
```

### Getting to Know the Participants
#### Differences in Gender
```{r}
library(ggplot2)
ggplot(camdata_clean, aes(x=gender, fill = gender))+ 
  geom_bar()+ 
  scale_fill_manual(values = c("blue4", "darkred"))+ 
  labs(title = "Gender Count in Survey Responses", x="Gender", y="Count", fill="Gender")
```

A notable gender disparity can be observed within the data set. The survey responses were predominantly from women with 112 respondents, whereas men represented a much smaller group with only 17 respondents. This stark difference in sample sizes raises important considerations while interpreting the findings in relation to gender. An important concept to this disparity may be due to voluntary response bias. The imbalance suggests that women may be more inclined to participate in surveys regarding spirituality, religiosity, or health in comparison to men. Notably, the small number of male respondents limits the ability to generalize findings across the general population. However, despite the imbalance, the data still provides valuable insights into how spirituality and religiosity influences health.

#### Marital Status
```{r}
ggplot(camdata_clean, aes(x=mrtlstus, fill = mrtlstus))+ 
  geom_bar()+ 
  scale_fill_manual(values = c("chartreuse3", "cornflowerblue", "brown", "aquamarine3"))+ 
  labs(title = "Marital Status Distribution in Survey Responses", x = "Status", y = "Count", fill="Marital Status")
```

The marital status of an individual can provide valuable insights into how personal relationships and social structures influence or interact with an individual's spiritual beliefs and practices. Married individuals generally experience higher levels of social support, which can positively influence their overall well-being. Spirituality and religiosity can play a role in strengthening marital bonds and providing emotional and practical support. This suggests that marital status impacts physical health outcomes. This suggests that marital status impacts physical health outcomes. The survey responses reflect that 53% respondents are married, 27% are single, and 19% are divorced, with 1 respondent being widowed. These percentages generally reflect marriage status within the UK in 2021 for single and married populations. Our sample contains a higher proportion of divorced participants and a lower proportion of widowed participants compared to the national average. Average age of respondents and accessibility of the survey to older populations may have affected the response rate of widowed participants. In general, we can assume that our sample is reflective of the marital status across the population. 

#### Denomination
```{r}
ggplot(camdata_clean, aes(x=denom, fill = denom))+ 
  geom_bar()+ 
  labs(title = "Religious Denomination Distribution", x = "Denomination", y = "Count", fill="Denomination")+ 
  theme(axis.text.x.bottom = element_text(angle = 45, hjust = 0.9))
```

When specifically analyzing the religious denominations of our participants, we see that 50 respondents were Protestant, followed by almost 30 stating they are in no denomination. The religious landscape in the UK has changed significantly in recent years. According to the 2021 Census for England and Wales, the largest group identifies as having no religion (46.2%), followed by Christians at 37.2%. Other religions include Muslims (6.5%), Hindus (1.7%), Sikhs (0.9%), Jews (0.5%), and Buddhists (0.5%)​. It is likely that the respondents were not entirely reflective of this, as they were located in a more XX area/neighborhood that more Protestants were being treated at. There is a notable trend in the population that identifies themselves as having no religion, or being atheist. Christianity remains the largest religious group in the UK, and in general a percentage of the population has become more religious in recent years since the Coronavirus pandemic.


#### How Religious or How Spiritual
```{r}
##install.packages(ggpubr)
library(ggpubr)

howreligious <- ggplot(camdata_clean, aes(x=relpersn, fill = relpersn))+  
  geom_bar()+ 
  scale_fill_manual(values = c("cadetblue2", "burlywood2", "brown2", "coral2"))+ 
  labs(title = "Self-Rated Religious Level", x = "Level of Religiousity", y = "Count", fill="Religiousity Scale")+
  theme(axis.text.x.bottom = element_text(size = 7.75))

howspiritual <- ggplot(camdata_clean, aes(x=sprtprsn, fill = sprtprsn))+ 
  geom_bar()+ 
  scale_fill_manual(values = c("cadetblue2", "burlywood2", "brown2", "coral2"))+ 
  labs(title = "Self-Rate Spiritual Level", x = "Level of Spirituality", y = "Count", fill="Spirituality Scale")+
  theme(axis.text.x.bottom = element_text(size = 7.75))

ggarrange(howreligious, howspiritual,
         ncol=1, nrow=2)
```
Interestingly, when asked to rate their levels to religiousness and spirituality, there are opposing trends: nearly 90 respondents see themselves as moderately to very spiritual, however approximately half the number see themselves as moderately to very spiritual. The participants seem to be less religious, and more spiritual. This is important to keep in mind when later analyzing their overall physical health. 

According to the Harvard Divinity School, Pew Research Center, and Psychology Today, the main differences between being religious and spiritual is the following: being religious is more about structured practices like observing holidays and following specific rites and ceremonies, following sacred texts, and recognizing the authority of religious leaders and institutions, whereas spirituality is more about an individual’s personal experience, rather than following established norms. In spirituality, there is significantly more flexibility. A person can explore their own being in a less structured manner, for example through meditation, yoga, mindfulness, personal reflection, and a connection with nature. One’s personal experiences may be drawn from various cultures and religions, and is focused on personal growth and their journey in life. It focuses more on being a human through universal concepts like love, compassion, and interconnectedness. 
While religion provides a structured path with community support and established doctrines, spirituality offers a more personal and flexible approach to exploring existential questions and the meaning of life.

### Analysis

#### Quantifying Health
We are interested in the intersection of religious and spiritual practices with physical and psychological health. Currently, all variables measuring health (both physical and psychological) are coded as factor variables, with levels "Not at all", "A little bit", "Moderately", "Quite a bit", and "Extremely" for rating the severity of these symptoms. In order to perform a detailed statistical analysis regarding the relationship between health and spirituality, we must first **quantify health symptoms** reported by these patients. This will allows us to measure "health" across different groups while still maintaining the integrity of survey responses. Quantifying health symptoms allows us the calculate the "average" health across groups and gives us a wider range of plots in order to assess our data. In order to quantify data, we must convert these factor variables into numerical variables. For the purposes of converting these variables, 0 will represent "Not at all", 1 will represent "A little bit", 2 will represent "Moderately", 3 will represent "Quite a bit", and 4 will represent "Extremely". A lower "health score" will correlate with a person reporting less severe symptoms, or a "healthier" respondent. 

```{r}
#physical health symptoms - appetite, nausea, weakness
  #appetite
camdata_clean$poor_appetite_num <- camdata_clean$poor_appetite
levels(camdata_clean$poor_appetite_num) <- c(0, 1, 2, 3, 4)
camdata_clean$poor_appetite_num <- as.numeric(as.character(camdata_clean$poor_appetite_num))

  #nausea
camdata_clean$nausea_num <- camdata_clean$nausea
levels(camdata_clean$nausea_num) <- c(0, 1, 2, 3, 4)
camdata_clean$nausea_num <- as.numeric(as.character(camdata_clean$nausea_num))

  #weakness
camdata_clean$weakness_num <- camdata_clean$weakness
levels(camdata_clean$weakness_num) <- c(0, 1, 2, 3, 4)
camdata_clean$weakness_num <- as.numeric(as.character(camdata_clean$weakness_num))

  #physical (combined)
camdata_clean$physical_health <- rowMeans(cbind(camdata_clean$poor_appetite_num, camdata_clean$nausea_num, camdata_clean$weakness_num), na.rm=T)


#psychological health symptoms - blameothers, lonely, mindwrong
  #blameothers
camdata_clean$blameothers_num <- camdata_clean$blameothers
levels(camdata_clean$blameothers_num) <- c(0, 1, 2, 3, 4)
camdata_clean$blameother_num <- as.numeric(as.character(camdata_clean$blameothers_num))

  #lonely
camdata_clean$lonely_num <- camdata_clean$lonely
levels(camdata_clean$lonely_num) <- c(0, 1, 2, 3, 4)
camdata_clean$lonely_num <- as.numeric(as.character(camdata_clean$lonely_num))

  #mindwrong
camdata_clean$mindwrong_num <- camdata_clean$mindwrong
levels(camdata_clean$mindwrong_num) <- c(0, 1, 2, 3, 4)
camdata_clean$mindwrong_num <- as.numeric(as.character(camdata_clean$mindwrong_num))

  #psychological (combined)
camdata_clean$psych_health <- rowMeans(cbind(camdata_clean$blameothers_num, camdata_clean$lonely_num, camdata_clean$mindwrong_num), na.rm=T)


#health (combined) - combining health scores of both physical and psychological symptoms
camdata_clean$health_total <- rowMeans(cbind(camdata_clean$physical_health, camdata_clean$psych_health), na.rm=T)
```

#### Physical Health vs. Psychological Health
Before delving into the relationship between spirituality and health, we should first measure whether there is a correlation between physical and psychological health. This will help us with our future analysis in comparing the physical and psychological health of different subgroups when assessing the impact of spiritual and religious practices. While physical and psychological symptoms are distinct, physical maladies can certainly lead to a lower quality of life and feelings of loneliness and helplessness. We predict at least a small correlation, if not a significant relationship between a person's physical and psychological health. 

```{r}
##physical vs psychological health
ggplot(camdata_clean, aes(x=physical_health, y=psych_health))+
  geom_point(position="jitter")+
  geom_smooth(color="blue4")+
  labs(x="Physical Health Score", y="Psychological Health Score", title="Psychological vs. Physical Health in Survey Responses")
```
[Analysis - they are correlated. we expect this relationship to persist in our further analysis when introducing spirituality and religion.]

#### Spiritual vs. Religious - Health Effects
Two of our variables assess the self-ranking of respondents on how religious and how spiritual they consider themselves to be. We used this as a benchmark to assess the "religiousness" and "spirituality" of respondents. Religion is defined as a personal set of institutionalized system of religious attitudes and beliefs, while spirituality defines an experience of connection to something larger than you. Religion in essence consists of spirituality, in fostering beliefs of a connection to a higher power, although religion is more structured and faith-based while spirituality is more personal. Both religiosity and spirituality involve faith in the world, in a higher power, and both are linked to positive coping mechanisms that bolster mental health.

In the interests of this study, we are focused on studying the relationship between religion, spirituality, and health outcomes. We utilize our respondents’ self-ranking of their own religiousness and spirituality as a benchmark to scale their level of faith. We strive to determine the relationship between spirituality and religiousness and severity of health symptoms in order to better understand the role of spiritual and religious practices in the field of medicine. 
##### Are Spiritual People Religious? Are Religious People Spiritual?
In order to properly define the relationship between spirituality and health, we must first define how our respondents view these two ideas of faith, and how the survey respondents categorize themselves within the religious and spiritual spheres.

```{r}
##religion vs. spiritual self-ranking
camdata_clean %>%
  filter(!is.na(relpersn), !is.na(sprtprsn))%>%
  ggplot(aes(x=relpersn, y=sprtprsn))+
    geom_point(position="jitter")+
    geom_smooth()+
    labs(x="Religious Self-Ranking", y="Spiritual Self-Ranking", title="The Relationship between Spirituality and Religiosity")
```
Assessing the relationship between a respondent’s religious vs. spiritual self-ranking reveals no significant correlation between the two values. As discussed before, the majority of respondents categorize themselves as spiritual, but not many considered themselves religious. This marks an important distinction between religiousness and spirituality in the perspectives of our sample, perhaps highlighting the personal freedom of spirituality in the face of more structured religious practices. In order to determine what exactly differs between religious and spiritual practice would require further analysis into our respondents’ daily faith practices and activities. In general, there seems to be no statistical correlation between religious and spiritual self-ranking. “Very religious” people always ranked themselves spiritual, but results differed for “very spiritual” respondents. 

##### Does Spirituality/Religiosity Predict Severity of Health Symptoms?  
Having established the distinction between religiosity and spirituality, and having assessed the demographic self-ranking of our respondent’s religiousness and spirituality, we can now analyze the relationship between religiosity/spiritual vs. severity of health symptoms. 

The severity of health symptoms (overall, physical, and psychological) and in turn the health scores of respondents in this survey, are as follows:
```{r}
#overall health score
camdata_clean %>%
  summarize(med_health_total=median(health_total, na.rm=T),
            avg_health_total=mean(health_total, na.rm=T))
## The average overall health score is 0.93, and the median overall health score is 0.83.
## The majority of respondents rate themselves as experiencing overall (physical and psychological) health symptoms between "Not at all" and "A little bit". 
```
```{r}
#physical
camdata_clean %>%
  summarize(med_phys_health=median(physical_health, na.rm=T),
            avg_phys_health=mean(physical_health, na.rm=T))
## The average physical health score is 0.87, and the median physical health score is 0.67.
```
```{r}
#psychological
camdata_clean %>%
  summarize(med_psych_health=median(psych_health, na.rm=T),
            avg_psych_health=mean(psych_health, na.rm=T))
## The average psychological health score is 0.99, and the median psychological health score is 0.67.
```
In summary, the average overall health score is 0.93, while the median overall health score is 0.83. The median physical and psychological health scores are the same at 0.67, but the average psychological score is higher than the average physical score. In general, the majority of respondents report low physical and psychological symptoms, averaging between a “Not at all” and “A little bit”. Respondents tended to rate their psychological symptoms a bit more severe than their physical symptoms. 

```{r}
##religiosity vs health
camdata_clean %>%
   filter(!is.na(relpersn))%>%
  ggplot(aes(x=relpersn, y=health_total))+
    geom_point(position="jitter")+
    labs(x="Religious Self-Ranking", y="Overall Health Score", title="Overall Health Score vs. Religious Self-Ranking")
```
[Analysis here]

```{r}
#spirituality vs health
camdata_clean %>%
   filter(!is.na(sprtprsn))%>%
  ggplot(aes(x=sprtprsn, y=health_total))+
    geom_point(position="jitter")+
    labs(x="Spiritual Self-Ranking", y="Overall Health Score", title="Overall Health Score vs. Spiritual Self-Ranking")
```
[Analysis here]

#### How Does Marital Status Affect Health Symptoms? 
As we discussed previously, the marital status of an individual can reflect an individual's spiritual beliefs and practices, providing insight into a person's emotional and physical well-being. Married individuals tend to place more value of spiritual and religious beliefs, which in turn can strengthen marital bonds and introduce central values within couples. As we assessed prior, our survey responses contains a good mix of single, married, and divorced respondents representative of UK populations. By assessing the relationship between marital status and health, we are given insight as to how one's emotional support system can predict health and well-being. 
```{r}
##marital status vs health
camdata_clean %>%
  ggplot(aes(x=mrtlstus, y=health_total, fill=mrtlstus))+
  geom_bar(stat="summary", fun="mean")+
  scale_fill_manual(values = c("chartreuse3", "cornflowerblue", "brown", "aquamarine3"))+ 
  labs(x="Marital Status", y="Average Health Score", title="Average Health Score by Marital Status")
```
Combining both physical and psychological health, marriage predicts a lower average health score (less severe health symptoms) in survey respondents. Health score was calculated numerically by converting “Not at all", "A little bit", and "Moderately", to "Quite a bit", and "Extremely” responses to a scale from 0 to 4. Married individuals, compared to single and divorced respondents, had a median score of around 0.5 (between “Not at all” and “A little bit”), while single and divorced respondents both had a median score around 1 (at the “A little bit range). Considering our previous analysis where the median overall health score was below 1, We find this to be notable. Single and divorced individuals rated their overall health worse than the average participant. Marital status, in this survey, can be used to predict the severity of the respondents ratings; married individuals were more likely to rate their overall health symptoms less severe. A further analysis into the differences in rating between physical and psychological symptoms is warranted. 

##### Marital Status and Health Symptoms - Physical vs. Psychological
We've established that marital status predicts the severity of health symptoms, with married individuals scoring much lower in their average health score. However, the measurement we used, the "average health score" factors in all physical and psychological symptoms we previously selected - poor appetite, nausea, weakness, feeling others are to blame, feelings of loneliness, and feeling that something is wrong with one's mind. While marriage provides a significant emotional support system that might alleviate feelings of loneliness and blame, marriage is not directly linked to helping physical symptoms like nausea and weakness. Comparing how marital status predicts the severity of physical vs. psychological symptoms can provide insight into how the religious and spiritual aspects of marriage and partnership affect health symptoms. 

```{r}
##marital status vs health (physical/psych)
physvsmrtl <- camdata_clean %>%
  ggplot(aes(x=mrtlstus, y=physical_health, fill=mrtlstus))+
  geom_bar(stat="summary", fun="mean")+
  scale_fill_manual(values = c("chartreuse3", "cornflowerblue", "brown", "aquamarine3"))+
  labs(x="Marital Status", y="Physical Health Score", title="Physical Health Score by Marital Status")

psychvsmrtl <- camdata_clean %>%
  ggplot(aes(x=mrtlstus, y=psych_health, fill=mrtlstus))+
  geom_bar(stat="summary", fun="mean")+
  scale_fill_manual(values = c("chartreuse3", "cornflowerblue", "brown", "aquamarine3"))+
  labs(x="Marital Status", y="Psychological Health Score", title="Psychological Health Score by Marital Status")

ggarrange(physvsmrtl, psychvsmrtl,
          ncol=2, nrow=1)
```
The physical and psychological health score trends follow that of the overall health score trend, in that married individuals tend to rank the severity of their health symptoms lower than single and divorced individuals. In this study, people on average rated their psychological symptoms as more severe than their physical health symptoms, and this trend is observed within all categories of marital status in comparing physical and psychological symptoms. Marital status cannot be used as a method to predict psychological health symptoms over physical health symptoms, and vice versa. Whether this is due to the strong correlation between physical and psychological health, or because the emotional support and sacred bond of marriage that alleviates psychological symptoms also plays a role in mediating physica symptoms, is unclear. Further analysis on how these respondents engage in spiritual and religious practices, and how these practices are tied into marital status, is needed to create a conclusion about how marital status is related to spirituality, and how the relationship between the two plays a role in addressing health concerns. 

### Conclusion
