---
title: "Unveiling the Spiritual Connection to Health"
author: "Mariela Mendez, Hannah Huang"
date: "2024-07-25"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

### Complementary and Alternative Medical (CAM) and Spirituality, Religiosity Survey
The CAM and Spirituality, Religiosity survey was given to chronically ill patients who were randomly chosen from a patient database at a private medical clinic in Southampton, UK. To be a participant for the survey, individuals were required to (1) have recent treatment within the last three years, (2) have a diagnosis of either migraines, irritable bowel, or chronic fatigue syndrome, and (3) have an age greater than 18. Six hundred questionnaires were originally mailed to the patients, yet there was a response rate of 18.5%, with 111 successfully completed questionnaires. The survey aims to explore the impact of spirituality on health. The spiritual topics range from spiritual experiences, personal values, to private religious practices and more. As for the health topics, these range from mindfulness practices to current health issues. The following data is sourced from the Association of Religion Data Archives, www.theARDA.com, and was collected by principal investigators Niko Kohls, Harald Walach, and George Lewish. The original data set consists of 254 variables and 111 participants, but for better comprehension, the ARDA decided to add variables and responses to include 258 total variables and 130 respondents. However, for the purposes of this analysis, the variable count has been reduced to 13 variables. While this is a much shorter data set, the chosen variables create a more direct approach. A concise data set facilitates clearer interpretation of results without one being overwhelmed by excessive data. 

### Motivation
Understanding the relationship between spirituality and health is an important aspect in well-being. In medical environments, especially those dominated by Western medicine, spirituality's impact on health is often disregarded. With this data set, investigators are able to explore how spirituality affects various sectors of health and overall well-being. Via analysis, patterns or correlations can be linked. One could uncover whether certain spiritual practices or beliefs are associated with better mental health, reduced stress, or improved coping mechanisms. In return, these patterns are relevant to healthcare providers or therapists looking to integrate effective strategies into their treatments. Notably, findings could influence policies related to holistic healthcare approaches that acknowledge the role of spirituality in patient well-being.

<!-- ```{r}
nrow(camdata)
ncol(camdata)
# Our data set consists of 130 responses to 258 variables!
``` -->

### Defining Variables
The original data set consists of 130 responses to 258 variables! Narrowing the focus for a more direct approach, 13 variables were chosen instead. Three categories were made in order to create a distinction between these chosen variables. The categories are as follows: spirituality/religious, health effects, and personal data. For the first category, the variable *belgdwth* refers to the survey question "I believe in a God who watches over me", *prvtpry* refers to "How often do you pray privately in place?", *relpersn* refers to "to what extent do you consider yourself a religious person?", and *sprtprsn* refers to "to what extent do you consider yourself a spiritual person?". The health effects category contains both physical and psychological effects. These variables are *DISTRS11*: poor appetite, *DISTRS23*: nausea or upset stomach, *DISTRS37*: feeling weak in parts of your body, *DISTRS04*: feelings others are to blame for most of your troubles, *DISTRS16*: feeling lonely, *DISTRS53*: the idea that something is wrong with your mind. As for the personal data category, the variables chosen are *gender*: female or male, *mrtlstus*: marital status (single, married, divorced,  or widowed), and *denom*: denomination. COME BACK TO EXPLAIN REASONING FOR SELECTING THESE VARIABLES!!

```{r}
#load data set
load("/Users/marielamendez-pimentel/Documents/EDA Notebook/Datasets/CAM Original Dataset.RData")
#create new name for data set
camdata <- Complementary_and_Alternative_Medical_CAM_and_Spirituality_Religiosity_Survey
```

### Cleaning The Data
We will create a new data frame incorporating our variables of interest.

```{r, results='hide'}
#gives variable names and location
names(camdata) 

#access to needed libraries
library(haven)
library(purrr)
library(knitr)

#variables are stored as labelled vectors in the original data set so it must be converted to a regular vector so commands like summary apply to the variable
camdata <- camdata %>% mutate(belgdwth = as_factor(belgdwth)) 
camdata <- camdata %>% mutate(prvtpry = as_factor(prvtpry)) 
camdata <- camdata %>% mutate(relpersn = as_factor(relpersn)) 
camdata <- camdata %>% mutate(sprtprsn = as_factor(sprtprsn)) 
camdata <- camdata %>% mutate(DISTRS04 = as_factor(DISTRS04)) 
camdata <- camdata %>% mutate(DISTRS11 = as_factor(DISTRS11)) 
camdata <- camdata %>% mutate(DISTRS16 = as_factor(DISTRS16)) 
camdata <- camdata %>% mutate(DISTRS23 = as_factor(DISTRS23)) 
camdata <- camdata %>% mutate(DISTRS37 = as_factor(DISTRS37)) 
camdata <- camdata %>% mutate(DISTRS53 = as_factor(DISTRS53)) 
camdata <- camdata %>% mutate(gender = as_factor(gender)) 
camdata <- camdata %>% mutate(mrtlstus = as_factor(mrtlstus))
camdata <- camdata %>% mutate(denom = as_factor(denom)) 
``` 

```{r, results='hide'}
#creates new dataset with only the desired variables and saves to working project
camdata_clean <-camdata[c(154,159,188,189,195,202,207,214,228,244,246,247,250)]
summary(camdata_clean)
save(camdata_clean, file = "/Users/marielamendez-pimentel/Documents/EDA Notebook/Datasets/camdata_clean.RData")
```

### Getting to Know the Participants
#### Gender
```{r}
library(ggplot2)
ggplot(camdata_clean, aes(x=gender, fill = gender)) + geom_bar() + scale_fill_manual(values = c("blue4", "darkred")) + labs(title = "Gender", x="Gender", y="Count")
```

[Analysis here]
#### Marital Status
```{r}
ggplot(camdata_clean, aes(x=mrtlstus, fill = mrtlstus)) + geom_bar() + scale_fill_manual(values = c("chartreuse3", "cornflowerblue", "brown", "aquamarine3")) + labs(title = "Marital Status", x = "Status", y = "Count" )
```

[Analysis here]

#### Denomination
```{r}
ggplot(camdata_clean, aes(x=denom, fill = denom)) + geom_bar() + labs(title = "Religious Denomination", x = "Denomination", y = "Count" ) + theme(axis.text.x.bottom = element_text(angle = 45, hjust = 0.9))
```

[Anaylis here]

#### How Religious or How Spiritual
```{r, figures-side, fig.show='hold', out.width="50%"}
ggplot(camdata_clean, aes(x=relpersn, fill = relpersn)) +  geom_bar() + scale_fill_manual(values = c("cadetblue2", "burlywood2", "brown2", "coral2")) + labs(title = "Religious Level", x = "Level of Religiousity", y = "Count") + theme(axis.text.x.bottom = element_text(size = 7.75))

ggplot(camdata_clean, aes(x=sprtprsn, fill = sprtprsn)) + geom_bar() + scale_fill_manual(values = c("cadetblue2", "burlywood2", "brown2", "coral2")) + labs(title = "Spiritual Level", x = "Level of Spirituality", y = "Count") + theme(axis.text.x.bottom = element_text(size = 7.75))
```

[Analysis here]

### Analysis

####Scatterplot
```{r}
attach(camdata_clean)
```


Plans: huge difference in gender (women 112> men 17)
most are not religious (50) but very spiritual (45)
90 nothing is wrong with mind
50 protestant
most are married
most never have private practice

want to create and analyze charts/graphs (maybe more)
spiritual person w/ psychological problems vs religious
spiritual person w/ physical problems vs religious
mrtstus effect on health problems 
belgdwth effect on health problems

