---
title: "EDA Notebook"
author: "Mariela Mendez, Hannah Huang"
date: "2024-07-21"
output: github_document
---
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

### Our Dataset and Research Question
Our dataset comes from a survey from the Associaton of Religion Data Archives. This survey contains measures for examining the effect of spirituality on health. (more information). 

```{r}
#load dataset to RStudio from GitHub
load("~/Downloads/EDA_Notebook-main/Datasets/CAM Original Dataset.RData")

#create new name for dataset
camdata <- Complementary_and_Alternative_Medical_CAM_and_Spirituality_Religiosity_Survey
```

#### Defining our data
Now that we've uploaded our dataset, we can determine how large our dataset is and what variables we can narrow our focus to study. 
```{r}
nrow(camdata)
ncol(camdata)
## Our dataset consists of 130 responses to 258 variables!
```


##### Cleaning our data
We will create a new dataframe incorporating our variables of interest.
```{r}
## Creating camdata_clean
camdata_clean <- select(camdata, c('belgdwth', 'prvtpry', 'relpersn', 'sprtprsn', 'DISTRS04', 'DISTRS11', 'DISTRS16', 'DISTRS23', 'DISTRS37', 'DISTRS53', 'gender', 'mrtlstus', 'denom'))
str(camdata_clean)

```

#### Analysis
```{r}


```
