---
title: "Unveiling the Spiritual Connection to Health"
author: "Mariela Mendez, Hannah Huang"
date: "2024-07-25"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

### Introduction & Motivation
The CAM and Spirituality, Religiosity dataset is from the Association of Religion Data Archives (ARDA) which studies the effect of spirituality on health based on various factors on chronically ill patients from a private medical clinic in Southampton, UK. Some of the correlations include spiritual and meditative practice, spiritual experiences, personal values, spiritual coping, and religious support. Topics for health measures include mindfulness when facing current problems. 

Since modern medicine is dominated by more Western techniques, contrasting the more spiritual focus of Eastern medicine, understanding the relationship between spirituality and health is an important aspect to consider that is often disregarded in medical environments today. This dataset explores whether certain spiritual practices or beliefs are associated with better mental and physical health, reduced stress, or improved coping mechanisms. What is found in this dataset can be relevant to current healthcare providers and therapists searching for alternative treatment methods and strategies outside normative medical practices. Notably, findings could influence policies related to holistic healthcare approaches that acknowledge the role of spirituality in patient well-being and target audiences that may not seem as traditional, including how self-care is taught in public school systems, how severely injured patients are treated, and how mental health is dealt with in the workplace. 


### About The Dataset
The principal investigators Niko Kohls, Harald Walach, and George Lewish sent out their questionnaires to a randomized population of patients who met the following three criteria: undergone treatment within a 3 year mark, older than the the age of 18, and had a diagnosis of either migraines, irritable bowel, or chronic fatigue syndrome. The respondents consisted of 71 cases with chronic fatigue (65%), 29 (27%) individuals with an irritable bowel syndrome as well as 9 (8%) patients suffering from migraines, totaling to 130 respondents.

Six hundred questionnaires were originally mailed to the patients. With a response rate of 18.5%, 111 questionnaires were successfully completed. The survey questions focused on the impact of patients' spirituality and their overall physical health. The spiritual topics range from spiritual experiences, personal values, to private religious practices and more. As for the health topics, these range from mindfulness practices to current health issues. The final dataset consisted of 258 total variables, from 254 initially, and 130 respondents for a better understanding of the subject matter. For our project’s more focused analysis, the variable count has been reduced to 13 variables. By concentrating on the factors that would best describe our sample and their relationship with religion/spiritual practices, such as marital status, religious/ spiritual self-ranking, and their physical and mental health, we developed a more direct approach towards understanding the correlation between spirituality and health. 

### Defining Variables
The original data set consists of 130 responses to 258 variables. To narrow our analysis to the factors that best describe respondents’ spiritual practices and health, 13 variables were chosen in order to rid of repetitive patterns. Within the 13 remaining variables, three categories were created in order to differentiate between them. The categories consist of: spirituality and/or religion, effects on an individual’s physical health, and demographic data. 

* For the **spiritual and religious** category, the following variables are as defined: 
    * *belgdwth* refers to the survey question "I believe in a God who watches over me",
    * *prvtpry* refers to "How often do you pray privately in place?",
    * *relpersn* refers to "To what extent do you consider yourself a religious person?"
    * *sprtprsn* refers to "To what extent do you consider yourself a spiritual person?". 
* The **health effects category** contains both physical and psychological effects. These variables are:
    * *DISTRS11*: poor appetite,
    * *DISTRS23*: nausea or upset stomach, 
    * *DISTRS37*: feeling weak in parts of your body, 
    * *DISTRS04*: feelings others are to blame for most of your troubles,
    * *DISTRS16*: feeling lonely,
    * *DISTRS53*: the idea that something is wrong with your mind. 
* As for the personal data category, the variables chosen are:
    * *gender*: female or male, 
    * *mrtlstus*: marital status (single, married, divorced,  or widowed), 
    * and *denom*: denomination. 


#### Rationale for Selecting Variables
Given the abundant amount of original variables, the elected variables were selected carefully. The variables *belgwth*, *prvtpry*, *relpersn*, and *sprtprsn* are variables that were chosen to determine the level to which a participant may be spiritual or religious, determining how strong the correlation between their spirituality and health would be: 

* The variable *belgwth* indirectly questions the participant if they believe in a greater existence and if the participant believes that greater existence is watching them. 
* The variable *prvtpry* aims to see how religious a participant actually is outside of church. Oftentimes, individuals want a certain image of themselves to be displayed to the public, but private practices are a different topic.
* The variables *relpersn* and *sprtprsn* are a self ranking and sets how the participant views themselves on a scale from "not at all" to "very". 

The next set of variables relate to a participant's current health issues. They can be further split into physical and psychological effects. 

* The physical symptom variables were poor appetite (*DISTRS11*), nausea or upset stomach (*DISTRS23*), and feelings of weakness in one's body (*DISTRS37*). These variables were selected as they directly related to the requirements to the survey study of having irritable bowel or chronic fatigue syndrome. While the third indicator was migraines, the statements related to them were not as direct as these. So for the sake of a focused analysis, irritable bowel and chronic fatigue syndrome were preferred. 
* The psychological health effects chosen were feelings others were to blame for most of one's troubles (*DISTRS04*), feelings of loneliness (*DISTRS16*), and feelings that something is wrong with one's mind (*DISTRS53*). *DISTRS04* was chosen to see if participants would blame their problems on an entity rather than themselves while *DISTRS16* and *DISTRS53* were elected to see if the level of religiosity or spirituality affected how lonely one was or if they felt something was genuinely wrong with them. 

The last set of variables were demographic variables to get to know the data set population. These variables include gender, marital status, and denomination. Ultimately, these variables were chosen to illuminate any findings on the health relationships between the spiritual and physical in hopes to offer a shorter but to some extent holistic approach. 

(While creating our new dataset, we renamed certain variables for clarity during data analysis as follows: 

* *belgdwth* to *godwatches*
* *DISTRS11* to *poor_appetite*,
* *DISTRS23* to *nausea*, 
* *DISTRS37* to *weakness*, 
* *DISTRS04* to *blameothers*,
* *DISTRS16* to *lonely*,
* *DISTRS53* to *mindwrong*.)


```{r}
#load data set
load("/Users/marielamendez-pimentel/Documents/EDA Notebook/Datasets/CAM Original Dataset.RData")
#create new name for data set
camdata <- Complementary_and_Alternative_Medical_CAM_and_Spirituality_Religiosity_Survey
```

```{r}
#library access
##install.packages(tidyverse)
##install.packages(haven)
##install.packages(purrr)
##install.packages(knitr)
library(tidyverse)
library(haven)
library(purrr)
library(knitr)

#variables are stored as labeled vectors in the original data set so it must be converted to a regular vector so commands like summary apply to the variable
#create new  dataset, camdata_clean
camdata_clean <- camdata %>% 
  mutate(belgdwth = as_factor(belgdwth), #mutate labeled vectors to factors
         prvtpry = as_factor(prvtpry),
         relpersn = as_factor(relpersn),
         sprtprsn = as_factor(sprtprsn),
         DISTRS04 = as_factor(DISTRS04),
         DISTRS11 = as_factor(DISTRS11),
         DISTRS16 = as_factor(DISTRS16),
         DISTRS23 = as_factor(DISTRS23),
         DISTRS37 = as_factor(DISTRS37),
         DISTRS53 = as_factor(DISTRS53),
         gender = as_factor(gender),
         mrtlstus = as_factor(mrtlstus),
         denom = as_factor(denom)
        ) %>%
  select(belgdwth, prvtpry, relpersn, sprtprsn, DISTRS04, DISTRS11, DISTRS16, DISTRS23, DISTRS37, DISTRS53, gender, mrtlstus, denom) %>% #select variables of interest
  rename(godwatches = belgdwth, #rename for clarity
         poor_appetite = DISTRS11,
         nausea = DISTRS23,
         weakness = DISTRS37,
         blameothers = DISTRS04,
         lonely = DISTRS16,
         mindwrong = DISTRS53)

# summary(camdata_clean)

#save updated dataframe to github
save(camdata_clean, file = "/Users/marielamendez-pimentel/Documents/EDA Notebook/Datasets/camdata_clean.RData")
```

### Getting to Know the Participants
#### Differences in Gender
```{r}
library(ggplot2)
ggplot(camdata_clean, aes(x=gender, fill = gender))+ 
  geom_bar()+ 
  scale_fill_manual(values = c("blue4", "darkred"))+ 
  labs(title = "Gender Count in Survey Responses", x="Gender", y="Count", fill="Gender")
```

A notable gender disparity can be observed within the data set. The survey responses were predominantly from women, who made up 112 of the responses, compared to the men that represented a much smaller percentage with only 17 respondents. One remaining respondent did not provide a direct answer to the question. This stark difference in gender raises important sample considerations, including voluntary response bias. The imbalance suggests that women are more likely to participate in surveys about spirituality, religiosity, or health than men, possibly because they are more involved in spiritual or religious practices and more willing to share their beliefs in surveys. Notably, the small number of male respondents limits the ability to generalize findings across the general population. However, despite the imbalance, the data still provides valuable insights into how spirituality and religiosity influences health in the human body and its nervous system. 

#### Marital Status
```{r}
ggplot(camdata_clean, aes(x=mrtlstus, fill = mrtlstus))+ 
  geom_bar()+ 
  scale_fill_manual(values = c("chartreuse3", "cornflowerblue", "brown", "aquamarine3"))+ 
  labs(title = "Marital Status Distribution in Survey Responses", x = "Status", y = "Count", fill="Marital Status")
```

Marital status provides insights into how personal relationships affect spiritual beliefs and practices. Married individuals typically have more social support, which enhances their well-being. Spirituality and religiosity can strengthen marital bonds, offering emotional and practical support, and thus positively impact physical health outcomes. The survey responses reflect that 53% of respondents are married, 27% are single, and 19% are divorced, with 1 respondent being widowed. These percentages generally reflect marriage status within the UK in 2021 for single and married populations. Our sample contains a higher proportion of divorced participants and a lower proportion of widowed participants compared to the national average. The average age of respondents and accessibility of the survey to older populations may have affected the response rate of widowed participants. In general, we can assume that our sample is relatively reflective of the married demographic across the population within the UK. 

#### Denomination
```{r}
ggplot(camdata_clean, aes(x=denom, fill = denom))+ 
  geom_bar()+ 
  labs(title = "Religious Denomination Distribution", x = "Denomination", y = "Count", fill="Denomination")+ 
  theme(axis.text.x.bottom = element_text(angle = 45, hjust = 0.9))
```

When specifically analyzing the religious denominations of our participants, we see that 50 respondents were Protestant, followed by nearly 30 individuals stating they are in no denomination. The religious landscape in the UK has changed significantly in recent years. According to the 2021 Census for England and Wales, the largest group identifies as having no religion (46.2%), followed by Christians at 37.2%. Other religions include Muslims (6.5%), Hindus (1.7%), Sikhs (0.9%), Jews (0.5%), and Buddhists (0.5%)​. Because the survey was only administered at one private medical clinic in Southampton, it is likely that the religious denomination distribution from respondents were not entirely reflective of that of the UK. It is probable that this was due to the clinic being located in an area with more Protestants. Additionally, we must also consider response bias, as those more involved in the religious and spiritual spheres of life are more inclined to participate in this survey. 

#### How Religious or How Spiritual
```{r}
##install.packages(ggpubr)
library(ggpubr)

howreligious <- ggplot(camdata_clean, aes(x=relpersn, fill = relpersn))+  
  geom_bar()+ 
  scale_fill_manual(values = c("cadetblue2", "burlywood2", "brown2", "coral2"))+ 
  labs(title = "Self-Rated Religious Ranking", x = "Level of Religiousity", y = "Count", fill="Religiousity Scale")+
  theme(axis.text.x.bottom = element_text(size = 7.75))

howspiritual <- ggplot(camdata_clean, aes(x=sprtprsn, fill = sprtprsn))+ 
  geom_bar()+ 
  scale_fill_manual(values = c("cadetblue2", "burlywood2", "brown2", "coral2"))+ 
  labs(title = "Self-Rated Spiritual Ranking", x = "Level of Spirituality", y = "Count", fill="Spirituality Scale")+
  theme(axis.text.x.bottom = element_text(size = 7.75))

ggarrange(howreligious, howspiritual,
         ncol=1, nrow=2)
```
Interestingly enough, when asked to rate their levels of religiousness and spirituality, responses reflected opposing trends: nearly 90 respondents see themselves as moderately to very spiritual, and approximately 40 see themselves as moderately to very religious. The participants view themselves as  less religious and more spiritual overall. This is important to keep in mind when later analyzing overall physical and psychological health. 

According to the Harvard Divinity School, Pew Research Center, and Psychology Today, the main differences between being religious and spiritual is the following: being religious is more about structured practices like observing holidays and following specific rites and ceremonies, following sacred texts, and recognizing the authority of religious leaders and institutions, whereas spirituality is more about an individual’s personal experience, rather than following established norms. In spirituality, there is significantly more flexibility. A person can explore their own being in a less structured manner, for example through meditation, yoga, mindfulness, personal reflection, and a connection with nature. One’s spiritual experiences may be drawn from various cultures and religions, and is focused on personal growth and their journey in life. Spirituality focused more on being a human through universal concepts like love, compassion, and interconnectedness. 

### Analysis

#### Quantifying Health
We are interested in exploring how religious and spiritual practices relate to physical and psychological health. Currently, health variables are coded as factor variables, with symptom severity measured in levels from "Not at all", "A little bit", and "Moderately", to "Quite a bit", and "Extremely." To conduct a detailed statistical analysis, we decided to **quantify health symptoms** by converting these factor variables into numerical values. This quantification allows us to measure "health" consistently across different groups, calculate average health scores, and generate various plots for data assessment, all while maintaining the integrity of survey responses.

```{r}
#physical health symptoms - appetite, nausea, weakness
  #appetite
camdata_clean$poor_appetite_num <- camdata_clean$poor_appetite
levels(camdata_clean$poor_appetite_num) <- c(0, 1, 2, 3, 4)
camdata_clean$poor_appetite_num <- as.numeric(as.character(camdata_clean$poor_appetite_num))

  #nausea
camdata_clean$nausea_num <- camdata_clean$nausea
levels(camdata_clean$nausea_num) <- c(0, 1, 2, 3, 4)
camdata_clean$nausea_num <- as.numeric(as.character(camdata_clean$nausea_num))

  #weakness
camdata_clean$weakness_num <- camdata_clean$weakness
levels(camdata_clean$weakness_num) <- c(0, 1, 2, 3, 4)
camdata_clean$weakness_num <- as.numeric(as.character(camdata_clean$weakness_num))

  #physical (combined)
camdata_clean$physical_health <- rowMeans(cbind(camdata_clean$poor_appetite_num, camdata_clean$nausea_num, camdata_clean$weakness_num), na.rm=T)


#psychological health symptoms - blameothers, lonely, mindwrong
  #blameothers
camdata_clean$blameothers_num <- camdata_clean$blameothers
levels(camdata_clean$blameothers_num) <- c(0, 1, 2, 3, 4)
camdata_clean$blameother_num <- as.numeric(as.character(camdata_clean$blameothers_num))

  #lonely
camdata_clean$lonely_num <- camdata_clean$lonely
levels(camdata_clean$lonely_num) <- c(0, 1, 2, 3, 4)
camdata_clean$lonely_num <- as.numeric(as.character(camdata_clean$lonely_num))

  #mindwrong
camdata_clean$mindwrong_num <- camdata_clean$mindwrong
levels(camdata_clean$mindwrong_num) <- c(0, 1, 2, 3, 4)
camdata_clean$mindwrong_num <- as.numeric(as.character(camdata_clean$mindwrong_num))

  #psychological (combined)
camdata_clean$psych_health <- rowMeans(cbind(camdata_clean$blameothers_num, camdata_clean$lonely_num, camdata_clean$mindwrong_num), na.rm=T)


#health (combined) - combining health scores of both physical and psychological symptoms
camdata_clean$health_total <- rowMeans(cbind(camdata_clean$physical_health, camdata_clean$psych_health), na.rm=T)
```

#### Physical Health vs. Psychological Health
Before delving into the relationship between spirituality and health, we should first determine if there is a correlation between physical and psychological health. Establishing this relationship is crucial for comparing the health of different subgroups when assessing the impact of spiritual and religious practices. Although physical and psychological symptoms are distinct, physical ailments can lead to lower quality of life and feelings of loneliness and helplessness, and vice versa. We anticipate finding at least a small, if not significant, correlation between a person's physical and psychological health. 

```{r}
##physical vs psychological health
ggplot(camdata_clean, aes(x=physical_health, y=psych_health))+
  geom_point(position="jitter")+
  geom_smooth(color="blue4")+
  labs(x="Physical Health Score", y="Psychological Health Score", title="Psychological vs. Physical Health in Survey Responses")
```
Upon further analysis, the plot demonstrates that the physical and psychological health severity are in fact correlated, however we cannot fully grasp how strong the correlation is due to only 130 responses out of an intended 600. A majority of survey participants have a low severity score, closer to zero on a range from zero to four, on psychological and physical health. The closer they are to zero, the healthier they are overall. In order to fully understand how strong the correlation is, data from more participants and different demographics, including more males and religious sectors would be needed. Regardless, knowing that there is a correlation between these two factors can help generalize medical solutions towards both physical and mental health in future analyses and in the healthcare industry. 

#### Spiritual vs. Religious - Health Effects
Two of our variables assess the self-ranking of respondents on how religious and how spiritual they consider themselves to be. We used this as a benchmark to assess the "religiousness" and "spirituality" of respondents. As earlier defined, religion follows more of an institutionalized system of attitudes and beliefs, while spirituality defines an experience of connection to something larger than you. However, both are linked to positive coping mechanisms that bolster mental, and thus, physical health.

##### Are Spiritual People Religious? Are Religious People Spiritual?
In order to properly define the relationship between spirituality and health, we must first define how our respondents view these two ideas of faith, and how the survey respondents categorize themselves within the religious and spiritual spheres.

```{r}
##religion vs. spiritual self-ranking
camdata_clean %>%
  filter(!is.na(relpersn), !is.na(sprtprsn))%>%
  ggplot(aes(x=relpersn, y=sprtprsn))+
    geom_point(position="jitter")+
    geom_smooth()+
    labs(x="Religious Self-Ranking", y="Spiritual Self-Ranking", title="The Relationship between Spirituality and Religiosity")
```
Assessing the relationship between a respondent’s religious vs. spiritual self-ranking reveals no significant correlation between the two values. As discussed before, the majority of respondents categorize themselves as spiritual, but not many considered themselves religious. This marks an important distinction between religiousness and spirituality in the perspectives of our sample, perhaps highlighting the personal freedom of spirituality in the face of more structured religious practices. In order to determine what exactly differs between religious and spiritual practice would require further analysis into our respondents’ daily faith practices and activities. In general, there seems to be no statistical correlation between religious and spiritual self-ranking. “Very religious” people always ranked themselves spiritual, but results differed for “very spiritual” respondents. 

##### Does Spirituality/Religiosity Predict Severity of Health Symptoms?  
Having established the distinction between religiosity and spirituality, and having assessed the demographic self-ranking of our respondent’s religiousness and spirituality, we can now analyze the relationship between religiosity/spiritual vs. severity of health symptoms. 

The severity of health symptoms, and in turn the health scores of respondents, are as follows:
```{r}
#overall health score
camdata_clean %>%
  summarize(med_health_total=median(health_total, na.rm=T),
            avg_health_total=mean(health_total, na.rm=T))
## The average overall health score is 0.93, and the median overall health score is 0.83.
## The majority of respondents rate themselves as experiencing overall (physical and psychological) health symptoms between "Not at all" and "A little bit". 
```
```{r}
#physical
camdata_clean %>%
  summarize(med_phys_health=median(physical_health, na.rm=T),
            avg_phys_health=mean(physical_health, na.rm=T))
## The average physical health score is 0.87, and the median physical health score is 0.67.
```
```{r}
#psychological
camdata_clean %>%
  summarize(med_psych_health=median(psych_health, na.rm=T),
            avg_psych_health=mean(psych_health, na.rm=T))
## The average psychological health score is 0.99, and the median psychological health score is 0.67.
```
In summary, the average overall health severity score is 0.93, while the median overall health severity score is 0.83. The average was taken using the mean (the sum of all responses over the total number of responses), and the median was calculated by finding the midpoint of responses. The median physical and psychological health severity scores were the same at 0.67, but the average psychological score is higher than the average physical score. In general, the majority of respondents report low physical and psychological symptoms, averaging less than 1 on a scale from 0 to 4. Respondents tended to rate their psychological symptoms slightly more severe than their physical symptoms.  

```{r}
##religiosity vs health
camdata_clean %>%
   filter(!is.na(relpersn))%>%
  ggplot(aes(x=relpersn, y=health_total))+
    geom_point(position="jitter")+
    labs(x="Religious Self-Ranking", y="Overall Health Score", title="Overall Health Score vs. Religious Self-Ranking")
```
When assessing the relationship between religious self-ranking and overall health severity score, there doesn’t seem to be a slight relationship between the two. Those with higher overall health severity tended to be less religious, and those who ranked themselves “Very religious” did not exceed an overall health severity score of 2, meaning their symptoms on average affected them only “a little bit.” As observed in our analysis of the severity of health outcomes, the majority of respondents’ overall health severity score did not exceed 1. This is reflected in our graph, with the majority of points clustered at the bottom. A larger survey, preferably with more respondents with higher severity of health symptoms, is needed to conclusively define the relationship between religiousness and health. 

```{r}
#spirituality vs health
camdata_clean %>%
   filter(!is.na(sprtprsn))%>%
  ggplot(aes(x=sprtprsn, y=health_total))+
    geom_point(position="jitter")+
    labs(x="Spiritual Self-Ranking", y="Overall Health Score", title="Overall Health Score vs. Spiritual Self-Ranking")
```
Compared to that of religion, there does not appear to be a relationship between personal spirituality and overall health severity. In the graph, we see our established trends of respondents ranking themselves highly spiritual, but having a lower overall health severity score less than 1. There is no observable trend in the level of spirituality of those who rated their overall health severity higher.

#### How Does Marital Status Affect Health Symptoms? 
As we discussed previously, the marital status of an individual can reflect an individual's spiritual beliefs and practices, providing insight into a person's emotional and physical well-being. Married individuals tend to place more value on spiritual and religious beliefs, which in turn can strengthen marital bonds and introduce central values within couples. By assessing the relationship between marital status and health, we are given insight as to how one's emotional support system can predict health and well-being. 
```{r}
##marital status vs health
camdata_clean %>%
  ggplot(aes(x=mrtlstus, y=health_total, fill=mrtlstus))+
  geom_bar(stat="summary", fun="mean")+
  scale_fill_manual(values = c("chartreuse3", "cornflowerblue", "brown", "aquamarine3"))+ 
  labs(x="Marital Status", y="Average Health Score", title="Average Health Score by Marital Status")
```
Combining both physical and psychological health, marriage predicts a lower average health severity score with less severe health symptoms in survey respondents. Married individuals, compared to single and divorced respondents, had a median severity score of around 0.5 (between “Not at all” and “A little bit”), while single and divorced respondents both had a median score around 1 (at the “A little bit” range). Considering our previous analysis where the median overall health severity score was below 1, this result is quite notable. Single and divorced individuals rated their overall health worse than the average participant. Marital status, in this survey, can be used to predict the severity of the respondents ratings, as married individuals were more likely to rate their overall health symptoms less severe. A further analysis into the differences in rating between physical and psychological symptoms is warranted. 

##### Marital Status and Health Symptoms - Physical vs. Psychological
We've established that marital status predicts the severity of health symptoms, with married individuals scoring much lower in their average health severity score. However, the measurement we used, the "average health severity score," factors in all physical and psychological symptoms we previously selected – poor appetite, nausea, weakness, feeling others are to blame, feelings of loneliness, and feeling that something is wrong with one's mind. While marriage provides a significant emotional support system that might alleviate feelings of loneliness and blame, marriage is not directly linked to helping physical symptoms like nausea and weakness. Comparing how marital status predicts the severity of physical vs. psychological symptoms can provide insight into how the religious and spiritual aspects of marriage and partnership also impact health symptoms. 

```{r}
##marital status vs health (physical/psych)
physvsmrtl <- camdata_clean %>%
  ggplot(aes(x=mrtlstus, y=physical_health, fill=mrtlstus))+
  geom_bar(stat="summary", fun="mean")+
  scale_fill_manual(values = c("chartreuse3", "cornflowerblue", "brown", "aquamarine3"))+
  labs(x="Marital Status", y="Physical Health Score", title="Physical Health Score by Marital Status")

psychvsmrtl <- camdata_clean %>%
  ggplot(aes(x=mrtlstus, y=psych_health, fill=mrtlstus))+
  geom_bar(stat="summary", fun="mean")+
  scale_fill_manual(values = c("chartreuse3", "cornflowerblue", "brown", "aquamarine3"))+
  labs(x="Marital Status", y="Psychological Health Score", title="Psychological Health Score by Marital Status")

ggarrange(physvsmrtl, psychvsmrtl,
          ncol=2, nrow=1)
```
The trends in physical and psychological health severity scores mirror the overall health severity scores, with married individuals generally reporting less severe health symptoms compared to single and divorced individuals. On average, respondents rated their psychological symptoms as more severe than their physical symptoms, a trend consistent across all marital status categories. However, marital status alone cannot predict psychological health over physical health or vice versa. This could be due to the strong correlation between physical and psychological health or the emotional support provided by marriage, which may alleviate both types of symptoms. Further research is needed to explore how spiritual and religious practices intersect with marital status and their combined effects on health outcomes.

### Conclusion
After narrowing the dataset down from 254 to 13 variables that describe the spiritual, religious and health-related factors among survey respondents, a few key takeaways have come to light. Comparing health symptoms amongst different demographics revealed that the majority of respondents reported low levels of both physical and psychological symptoms. Physical and psychological health are correlated, contrary to the current medical standard of treating disease over illness. This highlights the interconnected nature of overall health and how this concept of holistic treatment should be implemented into modern medicine. Additionally, there is no significant correlation between one’s level of religiosity and spirituality. Survey respondents tended to rank themselves as more spiritual and less religious. It remains important to consider the distinction between the two while implementing their practices in the process of healing. Neither religiosity nor spirituality seemed to demonstrate a significant impact on the health of the respondents. However, they can positively influence overall health through development of coping techniques and emotional support. Married individuals reported better health outcomes compared to single or divorced individuals, suggesting that emotional and social support in marriage helps reduce health symptoms. Considering the interplay between marriage and religious/spiritual practice, this finding supports the idea that demonstrating faith can contribute to alleviating the symptoms of chronically ill patients. It is clear that social factors that affect an individual’s mental strength can play a larger role in physical health than expected. 
These takeaways are significant because they emphasize the need for physicians and caretakers to gain a more holistic view of their patients in treating both physical and mental health. In today's Western approach to medicine, treatments are often direct, focusing on prescribing medications and expecting quick recoveries. However, this approach may overlook the significant impact of spiritual practice on mental health and physical healing. This dataset suggests that weaker mental health can hinder recovery, highlighting the importance of addressing both mental and physical health in medical care. Introducing faith-based practice may serve as an alternative in treating medical ailments, especially those affecting overall quality of life through chronic pain and illness. In addition, seeing the direct, positive effects on spirituality and religion on one’s overall health through survey responses, this amplifies the need for institutions and establishments to accommodate those who routinely partake in their religious practices, for example those who pray on a daily basis. If one is in touch with their culture and spirituality and are suddenly stripped of their routine and comfort in prayer or meditation, this may take a negative toll on both their mental and physical health. On the other hand, for those who may not have much experience in spirituality, programs can be offered to encourage patients to turn their nervous, stressed energy into hoping for positive change in their lives, resulting in a healthier mindset that will eventually lead to a more successful recovery. Further research with more participants and variables is needed to pinpoint specific goals and direct efforts in the effort to effectively integrate faith-based practices in the process of healing.

