---
title: "Unveiling the Spiritual Connection to Health"
author: "Mariela Mendez, Hannah Huang"
date: "2024-07-25"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

### Introduction & Motivation
In the context of modern medicine, which predominantly employs Western techniques, there is a growing need to consider the contrasting spiritual approaches found in alternative medicine. The integration of spirituality into healthcare is often overlooked in contemporary medical practices. This dataset aims to investigate whether specific spiritual practices or beliefs correlate with enhanced mental and physical health, reduced stress levels, or improved coping mechanisms. The insights derived from this dataset are highly relevant for healthcare professionals and therapists seeking  alternative treatment methods and strategies outside normative medical practices. The findings could inform holistic healthcare policies that integrate spirituality into patient care, influence educational programs on self-care in public schools, reshape treatment strategies for severely injured patients, and enhance mental health approaches in workplace settings. By acknowledging the role of spirituality in overall well-being, this research has the potential to shape more inclusive and comprehensive healthcare practices. 

### Dataset Collection Methodology: Survey Questionaire
The CAM and Spirituality, Religiosity dataset, sourced from the Association of Religion Data Archives (ARDA), examines the impact of spirituality on health among chronically ill patients from a private medical clinic in Southampton, UK. This dataset encompasses various dimensions of spirituality, including spiritual and meditative practices, personal values, spiritual coping strategies, and religious support. Additionally, it explores health measures related to mindfulness and overall coping when facing health challenges. 

The principal investigators Niko Kohls, Harald Walach, and George Lewish sent out their questionnaires to a randomized population of patients who met the following three criteria: undergone treatment within a 3 year mark, older than the the age of 18, and had a diagnosis of either migraines, irritable bowel, or chronic fatigue syndrome. The respondents consisted of 71 cases with chronic fatigue (65%), 29 (27%) individuals with an irritable bowel syndrome as well as 9 (8%) patients suffering from migraines, totaling to 130 respondents.

Six hundred questionnaires were originally mailed to the patients. With a response rate of 18.5%, 111 questionnaires were successfully completed. The survey questions focused on the impact of patients' spirituality and their overall physical health. The spiritual topics range from spiritual experiences, personal values, to private religious practices and more. As for the health topics, these range from mindfulness practices to current health issues. The final dataset consisted of 258 total variables, from 254 initially, and 130 respondents for a better understanding of the subject matter. For this project’s more focused analysis, the variable count has been reduced to 13 variables. By concentrating on the factors that would best describe a sample and their relationship with religion/spiritual practices, a more direct approach towards understanding the correlation between spirituality and health was developed. 

### Defining Chosen Variables 
The original data set consists of 130 responses to 258 variables. To narrow our analysis to the factors that best describe respondents’ spiritual practices and health, 13 variables were chosen in order to avoid repetitive patterns and create a more cohesive dataset. Within the 13 remaining variables, three categories were created in order to differentiate between them. The categories consist of: spirituality and/or religion, effects on an individual’s physical health, and demographic data. 

* For the **spiritual and religious** category, the following variables are as defined: 
    * *belgdwth* refers to the survey question "I believe in a God who watches over me",
    * *prvtpry* refers to "How often do you pray privately in place?",
    * *relpersn* refers to "To what extent do you consider yourself a religious person?"
    * *sprtprsn* refers to "To what extent do you consider yourself a spiritual person?". 
* The **health effects category** contains both physical and psychological effects. These variables are:
    * *DISTRS11*: poor appetite,
    * *DISTRS23*: nausea or upset stomach, 
    * *DISTRS37*: feeling weak in parts of your body, 
    * *DISTRS04*: feelings others are to blame for most of your troubles,
    * *DISTRS16*: feeling lonely,
    * *DISTRS53*: the idea that something is wrong with your mind. 
* As for the **personal data category**, the variables chosen are:
    * *gender*: female or male, 
    * *mrtlstus*: marital status (single, married, divorced,  or widowed), 
    * and *denom*: denomination. 

#### Rationale for Selecting Variables
Given the abundant amount of original variables, the elected variables were selected carefully. The variables *belgwth*, *prvtpry*, *relpersn*, and *sprtprsn* are variables that were chosen to determine the level to which a participant may be spiritual or religious, determining how strong the correlation between their spirituality and health would be: 

* The variable *belgwth* indirectly questions the participant if they believe in a greater existence and if the participant believes that greater existence is watching them. 
* The variable *prvtpry* aims to see how religious a participant actually is outside of church. Oftentimes, individuals want a certain image of themselves to be displayed to the public, but private practices are a different topic.
* The variables *relpersn* and *sprtprsn* are a self ranking and sets how the participant views themselves on a scale from "not at all" to "very". 

The next set of variables relate to a participant's current health issues. They can be further split into physical and psychological effects. 

* The physical symptom variables were poor appetite (*DISTRS11*), nausea or upset stomach (*DISTRS23*), and feelings of weakness in one's body (*DISTRS37*). These variables were selected as they directly related to the requirements to the survey study of having irritable bowel or chronic fatigue syndrome. While the third indicator was migraines, the statements related to them were not as direct as these. So for the sake of a focused analysis, irritable bowel and chronic fatigue syndrome were preferred. 
* The psychological health effects chosen were feelings others were to blame for most of one's troubles (*DISTRS04*), feelings of loneliness (*DISTRS16*), and feelings that something is wrong with one's mind (*DISTRS53*). *DISTRS04* was chosen to see if participants would blame their problems on an entity rather than themselves while *DISTRS16* and *DISTRS53* were elected to see if the level of religiosity or spirituality affected how lonely one was or if they felt something was genuinely wrong with them. 

The last set of variables were demographic variables to get to know the data set population. These variables include gender, marital status, and denomination. Ultimately, these variables were chosen to illuminate any findings on the health relationships between the spiritual and physical in hopes to offer a shorter but to some extent holistic approach. 

Note: While creating our new dataset, certain variables were renamed for clarity during data analysis as follows: 

* *belgdwth* to *godwatches*
* *DISTRS11* to *poor_appetite*,
* *DISTRS23* to *nausea*, 
* *DISTRS37* to *weakness*, 
* *DISTRS04* to *blameothers*,
* *DISTRS16* to *lonely*,
* *DISTRS53* to *mindwrong*.)

```{r}
#load data set
load("/Users/marielamendez-pimentel/Documents/EDA Notebook/Datasets/CAM Original Dataset.RData")
#create new name for data set
camdata <- Complementary_and_Alternative_Medical_CAM_and_Spirituality_Religiosity_Survey
```

```{r}
#library access
##install.packages(tidyverse)
##install.packages(haven)
##install.packages(purrr)
##install.packages(knitr)
library(tidyverse)
library(haven)
library(purrr)
library(knitr)

#variables are stored as labeled vectors in the original data set so it must be converted to a regular vector so commands like summary apply to the variable
#create new  dataset, camdata_clean
camdata_clean <- camdata %>% 
  mutate(belgdwth = as_factor(belgdwth), #mutate labeled vectors to factors
         prvtpry = as_factor(prvtpry),
         relpersn = as_factor(relpersn),
         sprtprsn = as_factor(sprtprsn),
         DISTRS04 = as_factor(DISTRS04),
         DISTRS11 = as_factor(DISTRS11),
         DISTRS16 = as_factor(DISTRS16),
         DISTRS23 = as_factor(DISTRS23),
         DISTRS37 = as_factor(DISTRS37),
         DISTRS53 = as_factor(DISTRS53),
         gender = as_factor(gender),
         mrtlstus = as_factor(mrtlstus),
         denom = as_factor(denom)
        ) %>%
  select(belgdwth, prvtpry, relpersn, sprtprsn, DISTRS04, DISTRS11, DISTRS16, DISTRS23, DISTRS37, DISTRS53, gender, mrtlstus, denom) %>% #select variables of interest
  rename(godwatches = belgdwth, #rename for clarity
         poor_appetite = DISTRS11,
         nausea = DISTRS23,
         weakness = DISTRS37,
         blameothers = DISTRS04,
         lonely = DISTRS16,
         mindwrong = DISTRS53)

# summary(camdata_clean)

#save updated dataframe to github
save(camdata_clean, file = "/Users/marielamendez-pimentel/Documents/EDA Notebook/Datasets/camdata_clean.RData")
```

### Getting to Know the Participants
Before analyzing the data within this survey, we must first familiarize ourselve with the respondents. This will give us insight into the demographic of our sample and whether our results can be generalized to the entire population. We focus on three different demographic categories: gender, marital status, and religious denomination.

#### Gender Disparity
```{r}
library(ggplot2)
ggplot(camdata_clean, aes(x=gender, fill = gender))+ 
  geom_bar()+ 
  scale_fill_manual(values = c("blue4", "darkred"))+ 
  labs(title = "Gender Count in Survey Responses", x="Gender", y="Count", fill="Gender")
```

The dataset reveals a significant gender disparity, with a predominance of female respondents. Out of 130 total responses, 112 are from women, while only 17 are from men. One respondent did not specify their gender. This notable imbalance raises important considerations regarding sample representation and potential response bias. The disproportion suggests that women are more likely to engage in surveys related to spirituality, religiosity, or health. This could be attributed to higher involvement in spiritual or religious practices among women, as well as a greater willingness to disclose personal beliefs and health experiences in surveys. The limited number of male respondents constrains the ability to generalize findings to the broader population. Despite this imbalance, the data provides valuable insights into the relationship between spirituality, religiosity, and health. The findings contribute to understanding how these factors influence well-being, particularly in the context of the human body and its nervous system.

#### Marital Status and Its Implications
```{r}
ggplot(camdata_clean, aes(x=mrtlstus, fill = mrtlstus))+ 
  geom_bar()+ 
  scale_fill_manual(values = c("chartreuse3", "cornflowerblue", "brown", "aquamarine3"))+ 
  labs(title = "Marital Status Distribution in Survey Responses", x = "Status", y = "Count", fill="Marital Status")
```

Marital status offers valuable insights into how personal relationships influence spiritual beliefs and practices. Typically, married individuals benefit from enhanced social support, which can positively affect their overall well-being. Spirituality and religiosity often play a role in strengthening marital bonds, providing both emotional and practical support, which can, in turn, improve physical health outcomes. According to the survey, 53% of respondents are married, 27% are single, 19% are divorced, and 1% are widowed. These proportions are generally consistent with the marriage status distribution in the UK as of 2021 for single and married populations. However, our sample shows a higher percentage of divorced individuals and a lower percentage of widowed individuals compared to national averages. This discrepancy may be attributed to the average age of respondents and the accessibility of the survey, which could have influenced the representation of widowed participants. Overall, while the sample is somewhat skewed, it still provides a reasonable reflection of the married demographic within the UK population. The findings highlight the significant role that marital status can play in shaping spiritual practices and their associated health benefits.

#### Religious Denominations of Participants
```{r}
ggplot(camdata_clean, aes(x=denom, fill = denom))+ 
  geom_bar()+ 
  labs(title = "Religious Denomination Distribution", x = "Denomination", y = "Count", fill="Denomination")+ 
  theme(axis.text.x.bottom = element_text(angle = 45, hjust = 0.9))
```

In analyzing the religious affiliations of our survey participants, it was discovered that 50 respondents identified as Protestant, while nearly 30 participants reported no religious affiliation. This distribution contrasts with broader religious trends in the UK. According to the 2021 Census for England and Wales, the largest demographic identifies as having no religion (46.2%), followed by Christians at 37.2%. Other religious affiliations include Muslims (6.5%), Hindus (1.7%), Sikhs (0.9%), Jews (0.5%), and Buddhists (0.5%).

The survey, conducted at a single private medical clinic in Southampton, may not fully represent the national religious landscape. The local demographics and the clinic's geographical location, which may have a higher concentration of Protestants, could have influenced the respondent distribution. Furthermore, response bias should be considered, as individuals more engaged in religious and spiritual practices might have been more inclined to participate in the survey.

#### Levels of Religiousness and Spirituality
```{r}
##install.packages(ggpubr)
library(ggpubr)

howreligious <- ggplot(camdata_clean, aes(x=relpersn, fill = relpersn))+  
  geom_bar()+ 
  scale_fill_manual(values = c("cadetblue2", "burlywood2", "brown2", "coral2"))+ 
  labs(title = "Self-Rated Religious Ranking", x = "Level of Religiousity", y = "Count", fill="Religiousity Scale")+
  theme(axis.text.x.bottom = element_text(size = 7.75))

howspiritual <- ggplot(camdata_clean, aes(x=sprtprsn, fill = sprtprsn))+ 
  geom_bar()+ 
  scale_fill_manual(values = c("cadetblue2", "burlywood2", "brown2", "coral2"))+ 
  labs(title = "Self-Rated Spiritual Ranking", x = "Level of Spirituality", y = "Count", fill="Spirituality Scale")+
  theme(axis.text.x.bottom = element_text(size = 7.75))

ggarrange(howreligious, howspiritual,
         ncol=1, nrow=2)
```

The survey results reveal a notable distinction between participants' self-perceived levels of spirituality and religiosity. Approximately 90 respondents identified themselves as moderately to very spiritual, while around 40 described themselves as moderately to very religious. This trend indicates that participants generally view themselves as more spiritual than religious. This distinction is significant for subsequent analyses of physical and psychological health, as it highlights different dimensions of personal belief systems. According to sources such as Harvard Divinity School, Pew Research Center, and Psychology Today, the key differences between religiosity and spirituality are as follows:

* Religiosity typically involves structured practices such as observing religious holidays, adhering to specific rites and ceremonies, following sacred texts, and recognizing the authority of religious leaders and institutions.
* Spirituality, in contrast, emphasizes personal experience and individual exploration. It is characterized by flexibility and can include practices such as meditation, yoga, mindfulness, personal reflection, and a connection with nature. Spirituality often draws from various cultural and religious traditions and focuses on personal growth, universal values such as love, compassion, and interconnectedness, and one's journey in life.

Understanding these distinctions is crucial for interpreting how spirituality and religiosity may impact overall health and well-being. The emphasis on spirituality over religiosity among participants may influence their health outcomes and experiences in different ways.

### Insights and Interpretation

#### Quantifying Health
This study aims to explore the relationship between religious and spiritual practices and both physical and psychological health. In the dataset, health variables are coded as factor variables, with symptom severity categorized into levels: "Not at all," "A little bit," "Moderately," "Quite a bit," and "Extremely." To enable a detailed statistical analysis, health symptoms are quantified by converting these categorical variables into numerical values. This quantification facilitates consistent measurement of health across different respondent groups, allows for the calculation of average health scores, and supports the generation of various plots for data visualization and assessment. This approach preserves the integrity of the survey responses while providing a rigorous framework for analysis.

```{r}
#physical health symptoms - appetite, nausea, weakness
  #appetite
camdata_clean$poor_appetite_num <- camdata_clean$poor_appetite
levels(camdata_clean$poor_appetite_num) <- c(0, 1, 2, 3, 4)
camdata_clean$poor_appetite_num <- as.numeric(as.character(camdata_clean$poor_appetite_num))

  #nausea
camdata_clean$nausea_num <- camdata_clean$nausea
levels(camdata_clean$nausea_num) <- c(0, 1, 2, 3, 4)
camdata_clean$nausea_num <- as.numeric(as.character(camdata_clean$nausea_num))

  #weakness
camdata_clean$weakness_num <- camdata_clean$weakness
levels(camdata_clean$weakness_num) <- c(0, 1, 2, 3, 4)
camdata_clean$weakness_num <- as.numeric(as.character(camdata_clean$weakness_num))

  #physical (combined)
camdata_clean$physical_health <- rowMeans(cbind(camdata_clean$poor_appetite_num, camdata_clean$nausea_num, camdata_clean$weakness_num), na.rm=T)


#psychological health symptoms - blameothers, lonely, mindwrong
  #blameothers
camdata_clean$blameothers_num <- camdata_clean$blameothers
levels(camdata_clean$blameothers_num) <- c(0, 1, 2, 3, 4)
camdata_clean$blameother_num <- as.numeric(as.character(camdata_clean$blameothers_num))

  #lonely
camdata_clean$lonely_num <- camdata_clean$lonely
levels(camdata_clean$lonely_num) <- c(0, 1, 2, 3, 4)
camdata_clean$lonely_num <- as.numeric(as.character(camdata_clean$lonely_num))

  #mindwrong
camdata_clean$mindwrong_num <- camdata_clean$mindwrong
levels(camdata_clean$mindwrong_num) <- c(0, 1, 2, 3, 4)
camdata_clean$mindwrong_num <- as.numeric(as.character(camdata_clean$mindwrong_num))

  #psychological (combined)
camdata_clean$psych_health <- rowMeans(cbind(camdata_clean$blameothers_num, camdata_clean$lonely_num, camdata_clean$mindwrong_num), na.rm=T)


#health (combined) - combining health scores of both physical and psychological symptoms
camdata_clean$health_total <- rowMeans(cbind(camdata_clean$physical_health, camdata_clean$psych_health), na.rm=T)
```

#### Physical Health vs. Psychological Health
Before delving into the relationship between spirituality and health, it is essential to first establish whether a correlation exists between physical and psychological health.  Establishing this relationship is crucial for comparing the health of different subgroups when assessing the impact of spiritual and religious practices. Although physical and psychological symptoms are distinct, physical ailments can lead to lower quality of life and feelings of loneliness and helplessness, and vice versa. It is anticipated that there will be at least a modest, if not significant, correlation between physical and psychological health. This preliminary analysis will provide a foundation for assessing how spiritual and religious practices impact overall well-being.

```{r, warning=FALSE, message=FALSE}
##physical vs psychological health
ggplot(camdata_clean, aes(x=physical_health, y=psych_health))+
  geom_point(position="jitter")+
  geom_smooth(method = 'lm', color="blue4")+
  labs(x="Physical Health Score", y="Psychological Health Score", title="Psychological vs. Physical Health in Survey Responses")
```

Further analysis of the data indicates that there is indeed a correlation between physical and psychological health severity. However, the strength of this correlation is somewhat limited by the relatively small sample size of 130 responses, compared to the intended 600. The majority of participants reported low severity scores on both psychological and physical health, with scores tending towards zero on a scale from zero to four, suggesting a generally healthy population. To gain a more accurate understanding of the correlation's strength, it would be beneficial to analyze data from a larger and more diverse sample, including a higher number of male respondents and individuals from various religious backgrounds. Despite these limitations, the identification of a correlation between physical and psychological health is valuable. It provides a foundation for future analyses and can inform more integrated approaches in medical solutions that address both physical and mental health within the healthcare industry.

#### Spiritual vs. Religious - Health Effects

Two variables in the dataset assess respondents' self-ratings of their religiousness and spirituality. These self-ratings serve as benchmarks for evaluating the "religiousness" and "spirituality" of participants. As previously defined, religion typically involves an institutionalized system of beliefs and practices, whereas spirituality is characterized by a personal sense of connection to something greater than oneself. Despite these distinctions, both dimensions are associated with positive coping mechanisms that can enhance mental health, which in turn may improve physical health.

##### Are Spiritual People Religious? Are Religious People Spiritual?

To accurately define the relationship between spirituality and health, it is essential to first understand how respondents perceive these concepts and how they classify themselves within the religious and spiritual domains. This understanding provides a foundational context for analyzing how these self-perceptions relate to health outcomes.

```{r, message=FALSE}
##religion vs. spiritual self-ranking
camdata_clean %>%
  filter(!is.na(relpersn), !is.na(sprtprsn))%>%
  ggplot(aes(x=relpersn, y=sprtprsn))+
    geom_point(position="jitter")+
    geom_smooth()+
    labs(x="Religious Self-Ranking", y="Spiritual Self-Ranking", title="The Relationship between Spirituality and Religiosity")
```

The analysis of the relationship between respondents’ self-ratings of religiousness and spirituality indicates no significant correlation between these two dimensions. As previously noted, the majority of respondents identify as spiritual rather than religious. This distinction underscores a divergence between religiousness and spirituality within the sample, potentially reflecting the personal flexibility associated with spirituality compared to the more structured nature of religious practices.

To fully understand the differences between religious and spiritual practices, further analysis is needed, particularly focusing on the daily faith practices and activities of respondents. Overall, the data reveals no statistical correlation between religious and spiritual self-ratings. While individuals who self-identified as "very religious" consistently also identified as spiritual, the same was not necessarily true for those who identified as "very spiritual."

##### Does Spirituality/Religiosity Predict Severity of Health Symptoms?  
Having clarified the distinction between religiosity and spirituality and assessed the demographic self-rankings of respondents in these areas, the next step is to analyze the relationship between religiosity/spirituality and the severity of health symptoms.

The severity of health symptoms, along with the corresponding health scores of respondents, is as follows:

##### Overlall Health
```{r, echo=FALSE}
library(dplyr)
library(knitr)

# Summarize health data
health_summary <- camdata_clean %>%
  summarize(
    med_health_total = median(health_total, na.rm = TRUE),
    avg_health_total = mean(health_total, na.rm = TRUE)
  )

# Display table
kable(health_summary)
```

The analysis of overall health scores reveals that the average score is 0.93, while the median score is 0.83. This indicates that most respondents have relatively low health symptom scores, with the majority reporting their overall (physical and psychological) health symptoms as falling between the categories of "Not at all" and "A little bit."

The average score of 0.93 suggests that, on average, respondents experience mild symptoms. However, the median score of 0.83 is slightly lower, indicating that more than half of the respondents report experiencing even fewer symptoms. The concentration of responses in the lower end of the severity scale reflects a generally positive health status among the sample population. This distribution provides a baseline understanding of the respondents' health conditions, which is crucial for evaluating how religiosity and spirituality may correlate with their physical and psychological well-being.

##### Physical Scores
```{r, echo=FALSE}
library(dplyr)
library(knitr)

# Summarize physical health data
physical_health_summary <- camdata_clean %>%
  summarize(
    med_phys_health = median(physical_health, na.rm = TRUE),
    avg_phys_health = mean(physical_health, na.rm = TRUE)
  )

# Display table
kable(physical_health_summary)
```

The analysis of physical health scores shows an average score of 0.87 and a median score of 0.67. The average score of 0.87 indicates that, on average, respondents report mild to moderate physical health symptoms. In contrast, the median score of 0.67 suggests that more than half of the respondents experience even fewer symptoms, reflecting a generally lower severity of physical health issues for the majority of the sample. This discrepancy between the average and median scores highlights that while some respondents report relatively higher physical health symptoms, a significant portion experiences fewer or less severe symptoms. The lower median score demonstrates that the majority of respondents are closer to the lower end of the severity scale, suggesting a generally favorable physical health status within the sample. 

##### Psychological Health Scores

```{r, echo=FALSE}
library(dplyr)
library(knitr)

# Summarize psychological health data
psych_health_summary <- camdata_clean %>%
  summarize(
    med_psych_health = median(psych_health, na.rm = TRUE),
    avg_psych_health = mean(psych_health, na.rm = TRUE)
  )

# Display table
kable(psych_health_summary)
```
The analysis of psychological health scores reveals an average score of 0.99 and a median score of 0.67. The average score of 0.99 indicates that, on average, respondents report mild psychological symptoms. This average value provides insight into the general level of psychological distress experienced by the participants. In contrast, the median score of 0.67 suggests that more than half of the respondents experience psychological symptoms that are less severe than this midpoint value. The median represents the central point of the data, indicating that the majority of participants report relatively lower levels of psychological distress. The discrepancy between the average and median scores illustrates that while the average score is slightly higher, indicating some variability in symptom severity, the median score reflects that a significant portion of respondents experience minimal distress. This variation suggests that, overall, the psychological health of the respondents is favorable, with many experiencing lower levels of psychological symptoms.

#### Health Severity Score Summary
In summary, the average overall health severity score is 0.93, while the median overall health severity score is 0.83. The average score, calculated as the mean of all responses, reflects the general level of health severity reported by participants. The median score, determined as the midpoint of the ordered responses, indicates that 50% of respondents experience health symptoms below this value, highlighting a slightly lower severity for the majority. The median scores for both physical and psychological health are the same at 0.67, which suggests that half of the respondents report physical and psychological symptoms of similar severity. However, the average psychological health score is higher at 0.99 compared to the average physical health score of 0.87. This indicates that, on average, respondents report slightly more severe psychological symptoms than physical symptoms. Overall, most respondents report low levels of both physical and psychological symptoms, with scores averaging less than 1 on a scale from 0 to 4. This pattern shows that while the majority experience mild symptoms, psychological symptoms are rated as somewhat more severe than physical symptoms.

##### Religious Self-Ranking and Health Severity
```{r, warning=FALSE}
##religiosity vs health
camdata_clean %>%
   filter(!is.na(relpersn))%>%
  ggplot(aes(x=relpersn, y=health_total))+
    geom_point(position="jitter")+
    labs(x="Religious Self-Ranking", y="Overall Health Score", title="Overall Health Score vs. Religious Self-Ranking")
```

An assessment of the relationship between religious self-ranking and overall health severity reveals no significant correlation. Individuals with higher overall health severity scores generally reported lower levels of religiosity. Specifically, those identifying as “Very religious” had overall health severity scores not exceeding 2, suggesting that their symptoms impacted their well-being only minimally. Data analysis indicates that the majority of respondents reported overall health severity scores below 1, as shown by the graph where most data points cluster at the lower end. To comprehensively understand the relationship between religiosity and health severity, a larger sample with a wider range of health severity levels is necessary.

```{r, warning=FALSE}
#spirituality vs health
camdata_clean %>%
   filter(!is.na(sprtprsn))%>%
  ggplot(aes(x=sprtprsn, y=health_total))+
    geom_point(position="jitter")+
    labs(x="Spiritual Self-Ranking", y="Overall Health Score", title="Overall Health Score vs. Spiritual Self-Ranking")
```

In contrast to religiosity, no significant relationship is observed between personal spirituality and overall health severity. The data shows that respondents who rated themselves as highly spiritual generally reported lower overall health severity scores, with most scores below 1. The graph indicates that there is no discernible trend linking higher levels of spirituality with increased health severity.

#### How Does Marital Status Affect Health Symptoms? 

Marital status can significantly influence an individual's spiritual beliefs and practices, and thereby impact emotional and physical well-being. Marriage often reinforces spiritual and religious values, fostering emotional support and enhancing the quality of life. Understanding the relationship between marital status and health provides valuable insights into how the presence or absence of a supportive partner can affect overall health and well-being. Married individuals typically benefit from a robust emotional support system, which can contribute to better health outcomes by mitigating stress and providing a stable foundation for both physical and mental health. Conversely, those who are single, divorced, or widowed may experience different levels of support and thus face varying health challenges.By analyzing the relationship between marital status and health symptoms, it is possible to gain a deeper understanding of how personal relationships and emotional support systems influence health. This analysis helps in identifying how different marital statuses correlate with health symptoms, potentially guiding interventions and support strategies tailored to individuals' relational contexts.
```{r, warning=FALSE}
##marital status vs health
camdata_clean %>%
  ggplot(aes(x=mrtlstus, y=health_total, fill=mrtlstus))+
  geom_bar(stat="summary", fun="mean")+
  scale_fill_manual(values = c("chartreuse3", "cornflowerblue", "brown", "aquamarine3"))+ 
  labs(x="Marital Status", y="Average Health Score", title="Average Health Score by Marital Status")
```

Combining both physical and psychological health metrics, marital status emerges as a significant predictor of overall health severity. Married respondents reported a notably lower average health severity score, reflecting less severe health symptoms compared to their single and divorced counterparts. Specifically, married individuals had a median severity score of approximately 0.5, which falls between the “Not at all” and “A little bit” ranges. In contrast, single and divorced respondents both reported median scores around 1, corresponding to the “A little bit” range. Given that the median overall health severity score for the entire sample was below 1, these findings are particularly striking. Single and divorced individuals rated their overall health more severely than the average participant, suggesting that marital status plays a crucial role in health outcomes. The data indicates that married individuals are more likely to report less severe health symptoms, which implies that marital status can be a useful predictor of health severity. Further analysis is needed to explore the differences in ratings between physical and psychological symptoms, to better understand how marital status influences various dimensions of health.

##### Marital Status and Health Symptoms - Physical vs. Psychological

The analysis has demonstrated that marital status is a predictor of overall health severity, with married individuals consistently reporting lower average health severity scores compared to single and divorced respondents. This overall health severity score encompasses a range of symptoms, including poor appetite, nausea, weakness, feelings of blame, loneliness, and concerns about mental health. While marriage appears to significantly reduce the perception of psychological symptoms such as loneliness and feelings of blame—likely due to the emotional support provided by a marital partner—its impact on physical symptoms like nausea and weakness is less clear. The emotional and psychological benefits of marriage may not directly address physical health issues. To gain a comprehensive understanding of how marital status influences health, it is crucial to distinguish between the impacts on physical versus psychological symptoms. This comparison can shed light on how the religious and spiritual dimensions of marital relationships contribute to the management and experience of both physical and psychological health symptoms. Further analysis in this area will provide deeper insights into the multifaceted role of marital status in health outcomes.

```{r, message=FALSE, warning=FALSE}
##marital status vs health (physical/psych)
library(ggplot2)
library(gridExtra)  # For ggarrange function

# Plot for Physical Health vs Marital Status
physvsmrtl <- camdata_clean %>%
  ggplot(aes(x=mrtlstus, y=physical_health, fill=mrtlstus)) +
  geom_bar(stat="summary", fun="mean", width=0.7) +  # Adjust bar width
  scale_fill_manual(values = c("chartreuse3", "cornflowerblue", "brown", "aquamarine3")) +
  labs(x="Marital Status", y="Physical Health Score", title="Physical Health Score") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),  # Rotate x-axis labels
        axis.title = element_text(size = 14),
        plot.title = element_text(size = 16),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10))

# Plot for Psychological Health vs Marital Status
psychvsmrtl <- camdata_clean %>%
  ggplot(aes(x=mrtlstus, y=psych_health, fill=mrtlstus)) +
  geom_bar(stat="summary", fun="mean", width=0.7) +  # Adjust bar width
  scale_fill_manual(values = c("chartreuse3", "cornflowerblue", "brown", "aquamarine3")) +
  labs(x="Marital Status", y="Psychological Health Score", title="Psychological Health Score") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),  # Rotate x-axis labels
        axis.title = element_text(size = 14),
        plot.title = element_text(size = 16),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10))

# Arrange plots side by side
ggarrange(physvsmrtl, psychvsmrtl, ncol=2, nrow=1)
```
The patterns observed in physical and psychological health severity scores reflect the overall health severity trends, with married individuals consistently reporting less severe health symptoms compared to their single and divorced counterparts. On average, respondents indicated that their psychological symptoms were more severe than their physical symptoms, a trend that was consistent across all marital status categories. However, marital status alone does not predict whether psychological health is more significantly affected than physical health, or vice versa. This lack of distinct prediction may be attributed to the strong correlation between physical and psychological health, or the possibility that the emotional support offered by marriage mitigates both types of symptoms. To fully understand the interplay between marital status, spiritual and religious practices, and their combined effects on health outcomes, further research is necessary.

#### Examples of Religious/Spiritual Practice: Private Pray
Recognizing the personalized nature of spiritual care, we seek to define and provide specific examples of religious and spiritual care that improve patient outcomes to implement into healthcare and patient treatments. One such example is implementing instances of private pray into one's daily routine, specifically for those who already define as religious. Private pray can help create a space for self-examination and confession for individuals honestly express their challenges and their hopes. Studies have shown that frequent private prayer is linked to significant benefits for mental health issues. We aim to examine the relationship between frequency of private pray and psychological health within this study. 

```{r}
install.packages("paletteer")
library(paletteer)

camdata_clean %>%
  filter(!is.na(prvtpry),
         relpersn==3 | relpersn==4) %>%
  group_by(prvtpry) %>%
  summarize(psychhealth_avg=mean(psych_health)) %>%
  ggplot(aes(x=prvtpry, y=psychhealth_avg, fill=prvtpry))+
    geom_col()+
    scale_fill_paletteer_d("MexBrewer::Frida")+
    labs(x="Frequency of Private Pray", y="Psychological Health Score", title="Private Pray In Relation to Psychological Health", fill="Private Pray Frequency")+
  theme(plot.title = element_text(family = "serif", size = 16, face = "bold"),
        axis.title.x = element_text(family = "serif", size = 13, face = "italic"),
        axis.title.y = element_text(family = "serif", size = 13, face = "italic"),
        axis.text.x.bottom = element_text(
          family = "serif", size = 11,
          angle = 40, hjust = 0.9),
        axis.text.y = element_text(family = "serif", size = 11),
        legend.title = element_text(family = "serif", size = 12),
        legend.text = element_text(family = "serif", size = 9))
  
```
For sampled individuals who identify as religious, an increased frequency of private pray predicts better mental health. Compared to individuals who never pray or pray less than once a month, those who seek a connection with God through prayer weekly or daily tend to rate their psychological health lower, indicating lower feelings of loneliness and helplessness. This suggests that private prayer might serve as a practical application to be investigated within the context of healthcare. If trained properly, physicians and other providers provide expertise on spiritual and religious care in order to help patients navigate mental health concerns within the context of disease and illness. Additionally, a deeper examination as to why private pray is advantageous to mental health might provide insight into applications for non-religious pateints. 

### Conclusion
After narrowing the dataset down from 254 to 13 variables that describe the spiritual, religious and health-related factors among survey respondents, a few key takeaways have come to light. Comparing health symptoms amongst different demographics revealed that the majority of respondents reported low levels of both physical and psychological symptoms. Physical and psychological health are correlated, contrary to the current medical standard of treating disease over illness. This highlights the interconnected nature of overall health and how this concept of holistic treatment should be implemented into modern medicine. Additionally, there is no significant correlation between one’s level of religiosity and spirituality. Survey respondents tended to rank themselves as more spiritual and less religious. It remains important to consider the distinction between the two while implementing their practices in the process of healing. Neither religiosity nor spirituality seemed to demonstrate a significant impact on the health of the respondents. However, they can positively influence overall health through development of coping techniques and emotional support. Married individuals reported better health outcomes compared to single or divorced individuals, suggesting that emotional and social support in marriage helps reduce health symptoms. Considering the interplay between marriage and religious/spiritual practice, this finding supports the idea that demonstrating faith can contribute to alleviating the symptoms of chronically ill patients. It is clear that social factors that affect an individual’s mental strength can play a larger role in physical health than expected. 
These takeaways are significant because they emphasize the need for physicians and caretakers to gain a more holistic view of their patients in treating both physical and mental health. In today's Western approach to medicine, treatments are often direct, focusing on prescribing medications and expecting quick recoveries. However, this approach may overlook the significant impact of spiritual practice on mental health and physical healing. This dataset suggests that weaker mental health can hinder recovery, highlighting the importance of addressing both mental and physical health in medical care. Introducing faith-based practice may serve as an alternative in treating medical ailments, especially those affecting overall quality of life through chronic pain and illness. In addition, seeing the direct, positive effects on spirituality and religion on one’s overall health through survey responses, this amplifies the need for institutions and establishments to accommodate those who routinely partake in their religious practices, for example those who pray on a daily basis. If one is in touch with their culture and spirituality and are suddenly stripped of their routine and comfort in prayer or meditation, this may take a negative toll on both their mental and physical health. On the other hand, for those who may not have much experience in spirituality, programs can be offered to encourage patients to turn their nervous, stressed energy into hoping for positive change in their lives, resulting in a healthier mindset that will eventually lead to a more successful recovery. Further research with more participants and variables is needed to pinpoint specific goals and direct efforts in the effort to effectively integrate faith-based practices in the process of healing.

